---
title: "Yield Curve Inversion"
author: "Author: MFA - Stream A - Group 7"
#date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```

#Predicting a recession

Every so often, we hear warnings from commentators on the "inverted yield curve" and its predictive power with respect to recessions. An explainer what a [inverted yield curve is can be found here](https://www.reuters.com/article/us-usa-economy-yieldcurve-explainer/explainer-what-is-an-inverted-yield-curve-idUSKBN1O50GA). If you'd rather listen to something, here is a great podcast from [NPR on yield curve indicators](https://www.podbean.com/media/share/dir-4zgj9-6aefd11)

In addition, many articles and commentators think that, e.g., [*Yield curve inversion is viewed as a harbinger of recession*](https://www.bloomberg.com/news/articles/2019-08-14/u-k-yield-curve-inverts-for-first-time-since-financial-crisis). One can always doubt whether inversions are truly a harbinger of recessions, and [use the attached parable on yield curve inversions](https://twitter.com/5_min_macro/status/1161627360946511873).`

In our case we will look at US data and use the [FRED database](https://fred.stlouisfed.org/) to download historical yield curve rates, and plot the yield curves since 1999 to see when the yield curves flatten. If you want to know more, a very nice article that explains the [yield curve is and its inversion can be found here](https://fredblog.stlouisfed.org/2018/10/the-data-behind-the-fear-of-yield-curve-inversions/).

##Editing the data in preparation of the analysis

First, we will use the `tidyquant` package to download monthly rates for different durations. 

```{r get_rates, warning=FALSE}
# Get a list of FRED codes for US rates and US yield curve; choose monthly frequency
# to see, eg., the 3-month T-bill https://fred.stlouisfed.org/series/TB3MS
tickers <- c('TB3MS', # 3-month Treasury bill (or T-bill)
             'TB6MS', # 6-month
             'GS1',   # 1-year
             'GS2',   # 2-year, etc....
             'GS3',
             'GS5',
             'GS7',
             'GS10',
             'GS20',
             'GS30')  #.... all the way to the 30-year rate

# Turn  FRED codes to human readable variables
myvars <- c('3-Month Treasury Bill',
            '6-Month Treasury Bill',
            '1-Year Treasury Rate',
            '2-Year Treasury Rate',
            '3-Year Treasury Rate',
            '5-Year Treasury Rate',
            '7-Year Treasury Rate',
            '10-Year Treasury Rate',
            '20-Year Treasury Rate',
            '30-Year Treasury Rate')

maturity <- c('3m', '6m', '1y', '2y','3y','5y','7y','10y','20y','30y')

# by default R will sort these maturities alphabetically; but since we want
# to keep them in that exact order, we recast maturity as a factor 
# or categorical variable, with the levels defined as we want
maturity <- factor(maturity, levels = maturity)

# Create a lookup dataset
mylookup<-data.frame(symbol=tickers,var=myvars, maturity=maturity)
# Take a look:
mylookup %>% 
  knitr::kable()

df <- tickers %>% tidyquant::tq_get(get="economic.data", 
                   from="1960-01-01")   # start from January 1960

glimpse(df)
```

Our dataframe `df` has three columns (variables):

- `symbol`: the FRED database ticker symbol
- `date`: already a date object
- `price`: the actual yield on that date

The first thing would be to join this dataframe `df` with the dataframe `mylookup` so we have a more readable version of maturities, durations, etc.

```{r join_data, warning=FALSE}

yield_curve <-left_join(df,mylookup,by="symbol") 
```

## Plotting the yield curve

### Yields on US rates by duration since 1960

Lets start this chapter of by looking at the yields on US treasury rates since 1960. We will create a seperate graph for each security we want to study. 

```{r yield_curve_per_type, out.width="100%"}
yield_curve$var_f<-factor(yield_curve$var,levels=c("3-Month Treasury Bill","6-Month Treasury Bill","1-Year Treasury Rate","2-Year Treasury Rate","3-Year Treasury Rate","5-Year Treasury Rate","7-Year Treasury Rate","10-Year Treasury Rate","20-Year Treasury Rate","30-Year Treasury Rate"))

yield_curve %>%  
  mutate(maturity = fct_relevel(maturity, c("1y","10y","2y","20y","3m","3y","30y","5y","5m","7y"))) %>%
  ggplot(aes(x=date, 
             y=price, 
             color = factor(var_f))) + 
  geom_line() + 
  facet_wrap(.~factor(var_f), nrow = 5) +
  labs(title="Yields on U.S. Treasury rates since 1960", 
       x = NULL, 
       y="%", 
       caption="Source: St Louis Federal Reserve Economic Database (FRED)") + 
  theme_bw() + 
  theme(legend.position = "none")
```

### Monthly yields on US rates by duration since 1999 on a year-by-year basis

Then, lets look at the yield curve for US treasury securities on a year-by-year basis since 1999.

```{r yield_curve_per_year, out.width="100%"}
yield_curve %>% 
  mutate(facet_year = year(ymd(date)), 
         facet_month = month(ymd(date))) %>%
  filter(facet_year >= 1999) %>%
  ggplot(aes(x=maturity, 
             y=price,
             color = factor(facet_year),
             group = factor(facet_month))) +
  geom_line() +
  facet_wrap(~facet_year,
             nrow = 6) +
  labs(title="US Yield Curve",
       x="Maturity", 
       y="Yield (%)", 
       caption="Source: St Louis Federal Reserve Economic Database (FRED)") +
  theme_bw() + 
  theme(legend.position = "none")
```


### 3-month and 10-year yields since 1999

Now, lets study the evelution of 3-month and 10-year yields since 1999 a little closer.

According to [Wikipedia's list of recession in the United States](https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States), since 1999 there have been two recession in the US: between Mar 2001–Nov 2001 and between Dec 2007–June 2009. We are especially looking to find out if the yieldcurve flattened before these recessions. If that is the case, can a yield curve flattening really mean a recession is coming in the US? Since 1999, when did short-term (3 months) yield more than longer term (10 years) debt?

```{r yield_curve_3m_and_10y, out.width="100%"}
yield_curve %>% 
  filter(maturity == "3m" |  maturity=="10y") %>%
  mutate(year = year(ymd(date))) %>%
  filter(year >= 1999) %>%
  ggplot(aes(x=date, 
             y=price, 
             color = var_f)) +
  geom_line() +
  labs(title="Yields on 3-month and 10-year US Treasury rates since 1999", 
       x=NULL, 
       y="%", 
       caption="Source: St Louis Federal Reserve Economic Database (FRED)") + 
  theme_bw()+ 
  theme(legend.title = element_blank())
```

In the above graph, we observe that the yield for 3 month treasury bills has surpassed the yield for 10- year treasury notes only 3 times between 1999 and 2020. Note that this event indicates an inverse yield curve and is therefore preceded by the yield curve flattening. The two recessions in 2001 and 2007-2009 were clearly preceded by the flattening of the yieldcurve. The most recent time this flattening happenned was in 2019, as you are well aware, this was also followed by a recession. This final observation would indicate that investors anticipated the current recession, be it not in its current form. 

##Creating a plot that illustrates our findings

Besides calculating the spread (10year - 3months), there are a few more things we need to do to produce a final plot that clearly summarizes all the things we have found so far and illustrates what we read/listened to in the attached documents.

1. Setup data for US recessions 
2. Superimpose recessions as the grey areas in our plot
3. Plot the spread between 30 years and 3 months as a blue/red ribbon, based on whether the spread is positive (blue) or negative(red)

- For the first, the code below creates a dataframe with all US recessions since 1946

```{r setup_US-recessions, warning=FALSE}

# get US recession dates after 1946 from Wikipedia 
# https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States

recessions <- tibble(
  from = c("1948-11-01", "1953-07-01", "1957-08-01", "1960-04-01", "1969-12-01", "1973-11-01", "1980-01-01","1981-07-01", "1990-07-01", "2001-03-01", "2007-12-01"),  
  to = c("1949-10-01", "1954-05-01", "1958-04-01", "1961-02-01", "1970-11-01", "1975-03-01", "1980-07-01", "1982-11-01", "1991-03-01", "2001-11-01", "2009-06-01") 
  )  %>% 
  mutate(From = ymd(from), 
         To=ymd(to),
         duration_days = To-From) %>%
  filter(To >= as.Date("1960-01-01"))
```

We will do the second and third step in the code below and plot the graph.

```{r final graph, out.width="100%"}
yield_curve %>% 
  filter(maturity == "3m" |  maturity=="10y") %>%
  arrange(date) %>%
  mutate(diff = lead(price) - price ) %>% 
  filter(maturity == "3m") %>%
  ggplot(aes(x=date, 
             y=diff)) +
  geom_line(size=0.25) +
  geom_hline(yintercept=0) +
  labs(title="Yield Curve Inversion: 10-year minus 3-month U.S. Treasury rates", 
       x=NULL, 
       y="Difference (10-year-3-month) tield in %", 
       subtitle="Difference in % points, monthly averages. \r\nShaded areas correspond to recessions",
       caption="Source: FRED, Federal Reserve Banks of St. Louis") + 
  theme_bw()+ 
  theme(legend.title = element_blank()) + 
  geom_ribbon(aes(x = date,
                  ymin = 0,
                  ymax = if_else(diff<0, diff, 0),
                  alpha=0.5,
                  fill = "A")) + 
  geom_ribbon(aes(x = date,
                  ymin = 0,
                  ymax = if_else(diff>0, diff, 0),
                  alpha=0.5,
                  fill = "B")) +
  scale_fill_manual(values=c("red","blue"),
                    name=NULL) +
  geom_rect(data=recessions,
            inherit.aes=F,
            aes(xmin=From, xmax=To, ymin=-Inf, ymax=+Inf),
                fill='grey',
                alpha=0.6) +
  theme(legend.position = "none") +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y") +
  geom_rug(aes(colour=ifelse(diff>=0,">=0","<0")),
               sides="b",
               alpha=0.5) +
  scale_colour_manual(values=c("red","blue"), 
                      guide=FALSE) +
  theme(plot.title = element_text(face="bold"),
        plot.subtitle=element_text(face="italic", size = 7))
```