---
title: "Gross Domestic Product"
author: "Author: MFA - Stream A - Group 7"
#date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```

#GDP components over time and among countries

At the risk of oversimplifying things, the main components of gross domestic product, GDP are personal consumption (C), business investment (I), government spending (G) and net exports (exports - imports). More information about GDP and the different approaches in calculating it can be found at [Wikipedia GDP page](https://en.wikipedia.org/wiki/Gross_domestic_product).

The GDP data we will look at is from the [United Nations' National Accounts Main Aggregates Database](https://unstats.un.org/unsd/snaama/Downloads), which contains estimates of total GDP and its components for all countries from 1970 to today. We will look at how GDP and its components have changed over time, and compare different countries and how much each component contributes to that country's GDP. The file we will work with is [GDP and its breakdown at constant 2010 prices in US Dollars](http://unstats.un.org/unsd/amaapi/api/file/6).

Lets start by looking at the excel to see how it is organised.

```{r read_GDP_data}
UN_GDP_data  <-  read_excel(here::here("data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip
```

##Tidying the data

The first thing you we want to do is to tidy the data, as it is in wide format and we must make it into long, tidy format in order to use it. We also made sure all figures are expressed in billions, and renamed the indicators into something shorter.

```{r reshape_GDP_data}

tidy_GDP_data  <- UN_GDP_data %>%
  pivot_longer(cols = 4:51, names_to = "year", values_to = "Expenditure") %>%mutate(Expenditure = Expenditure/(10^9))
  

clean_GDP_Data <- tidy_GDP_data%>%
  mutate(IndicatorName = case_when(
      IndicatorName == "Final consumption expenditure" ~ "Total Consumption", 
      IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~ "Household Consumption", 
      IndicatorName == "General government final consumption expenditure" ~ "Gov Consumption", 
      IndicatorName == "Gross capital formation" ~ "Gross Capital Formation",   
      IndicatorName == "Gross fixed capital formation (including Acquisitions less disposals of valuables)" ~ "Gross Capital", 
      IndicatorName == "Exports of goods and services" ~ "Exports", 
      IndicatorName == "Imports of goods and services" ~ "Imports", 
      IndicatorName == "Gross Domestic Product (GDP)" ~ "GDP", 
      IndicatorName == "Agriculture, hunting, forestry, fishing (ISIC A-B)" ~ "Ag/Hunt/Forest/Fish", 
      IndicatorName == "Mining, Manufacturing, Utilities (ISIC C-E)" ~ "Min/Manu/Utility",
      IndicatorName == "Manufacturing (ISIC D)" ~ "Manufacturing",
      IndicatorName == "Construction (ISIC F)" ~ "Construction",
      IndicatorName == "Wholesale, retail trade, restaurants and hotels (ISIC G-H)" ~ "Wholesale/Retail/Hospitality",
      IndicatorName == "Transport, storage and communication (ISIC I)" ~ "Transport/Comm/Storage",
      IndicatorName == "Other Activities (ISIC J-P)" ~ "Other",
      IndicatorName == "Total Value Added" ~ "Total Value Added",
      IndicatorName == "Changes in inventories" ~ "Changes in inventories"))

clean_GDP_Data

# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")

```

##Plot for GDP components over time

Lets produce a plot that gives us more information on GDP components over time.

```{r gdp1, out.width="100%"}
US_India_Germany_GDP <- clean_GDP_Data%>%
  group_by(year)%>%
  select(Country, IndicatorName, year, Expenditure)%>%
  filter(Country %in% c("United States","India","Germany"))%>%
  filter(IndicatorName %in% c("Gross Capital Formation", "Exports","Gov Consumption", "Household Consumption", "Imports" ))

ggplot(US_India_Germany_GDP, aes(year, Expenditure, colour = IndicatorName, group = IndicatorName)) + geom_line() + labs (title = "GDP components over time", subtitle = "In constant 2010 USD", y = "Billions US$", x = "") + theme_bw(base_size = 10) + scale_x_discrete(breaks = c(1970, 1980, 1990, 2000, 2010)) + facet_wrap(~Country) + guides(color=guide_legend("Components of GDP")) + theme(axis.text.x = element_text(size = 6))
```

##Difference between sum of GDP components and GDP as included in the dataframe

Now, recall that GDP is the sum of Household Expenditure (Consumption *C*), Gross Capital Formation (business investment *I*), Government Expenditure (G) and Net Exports (exports - imports). Even though there is an indicator `Gross Domestic Product (GDP)` in the dataframe, we will now recalculate it given its components discussed above.

Lets also calculate the % difference between what you calculated as GDP and the GDP figure included in the dataframe.

```{r}
GDP_Calculation <- clean_GDP_Data %>%
  filter(IndicatorName == "Household Consumption" | IndicatorName == "Gross Capital Formation" | IndicatorName == "Gov Consumption" | IndicatorName == "Exports" | IndicatorName == "Imports" | IndicatorName == "GDP")

GDP_Calculation <-  spread(GDP_Calculation, IndicatorName, Expenditure)

GDP_Calculation <- GDP_Calculation %>%
  mutate(Calculated_GDP = `Household Consumption`+`Gross Capital Formation`+`Gov Consumption`+Exports-Imports)

GDP_Calculation <- GDP_Calculation %>%
  mutate(Percentage_Difference = (Calculated_GDP - GDP)/GDP)
```

##GDP components plot conclusion

Finally, Lets study the previous chart in terms of proportions and discuss the different dynamics among these three countries. 

In terms of government expenditure, Germany and United States have been spending more than India, as a proportion of their respective GDPs. This could be attributed to Germany and United States being wealthier countries than India, with governments that have a deeper coffer as well as the ability to raise more debt.

We can also see an increase in gross capital formation in India, in the years after 2000, while the same levels are observed for Germany and United States. This could be interpreted as India accelerating its growth, to catch up to the more developed countries like Germany and United States.

The one factor that is consistent across all three countries is that household expenditure always comprises the biggest proportion of GDP, with consumers' spending being a key driver of any economy.